Constraining the High Density Nuclear Symmetry Energy with Pions

### 2020-07-17 chapters/ch_results.tex
Added bills edits working on appendix figures

 The data taken has been compared with 7 commonly used transport theoretical codes. These codes took part in a large collaborative effort in order to standardize certain common elements of each codes \cite{theoryComp1,theoryComp2}. To reduce uncertainties in the pion production mechanism, each model used comparison project simulated nuclear matter in a box type simulation with fixed, well defined, initial conditions. Though these box calculations model an equilibrated system at finite density, which is  not replicated by experiment, the limit is theoretically clearly defined and provides a benchmark comparison between models where the solutions  yield values for the total number of produced pions and $\Delta$ that can be analytically calculated from the ansatz of  statistical equilibrium, providing a good benchmark. This allowed for each code to systematically go through the numerical treatment and details in each code such as initialization of nuclei, stability of the code, numerical handing of pauli-blocking, etc. After satisfying these benchmark calculations, each model was used with what the authors considered the most reliable input and model assumption.  codes were taken in their best configuration, as a result of these comparison projects, without any prior knowledge of the experimental data. We then simulated the 4 systems measured in the \spirit TPC at an impact parameter of \SI{3}{\femto\metre} at \SI{270}{\MeVA} beam energy. While the numerical treatments of each code are reasonably similar, each code differs considerably in their treatment of pion and $\Delta$ dynamics. Some codes contain modifications to how the pion behaves in nuclear matter, i.e. \emph{in-medium effects}, typically introduced by including a pion optical potential which describes the pion scattering and absorption. Some codes include the iso-scalar and the iso-vector delta potential, which are not very well constrained but can be important in the production of pions \cite{baoan_deltapotential,inmedPionKo,inmedPionFeng}. The set of 7 codes represent the best codes at the moment, which can simulate pions at these energies.

It is important to note that the interaction of pions and deltas with dense matter should not be viewed as a nasty detail that a "better" probe like nucleonic elliptical flow would avoid. One of the central questions regarding matter at twice saturation density concerns the importance of deltas and pions as important constituents of the matter at those densities \cite{pionNS,deltaNS,awayforward}. Questions of the role of delta's and pion at 2$\rho_o$ are strongly connect to these potentials. Calculations of pion production indicate that selected pionic observables such as the energy spectra and transverse momentum distributions can be used to constrain these densities \cite{cozmaPC}. Obtaining such constraints are a major objective of our current work, but such constraints are ready for discussion at the present time.

As will be the re-occurring theme, there is a large variation in the predicted observables, between theoretical codes; greater than the variation between different symmetry energy in a particular code. However, the remedy for this discrepancy is for the authors of these models to include the other physics that many of them have neglected. This includes a better description of delta interactions of matter, which is critical to reproducing the total pion yields, and also the momentum dependence of the isovector mean field potentials, which have a large impact on the shapes of the energy spectra. Though some of the elementary issues relevant to such transport theories, such as the Pauli principal and the incorporation of inelastic processes, have been addressed. The relevant physics include the mean field potentials for nucleons, pions and deltas. A primary focus is on the isovector potentials of which only the momentum independent potentials for nucleons have been varied in the calculations shown below. The momentum dependent nucleonic mean field potentials and the mean field potentials for delta's and pions are only included in a subset of calculations. Calculations have already identified that the overall pion yield are influenced by the mean field potentials for the delta's \cite{cozmaPC}, which may not be surprising since a similar sensitivity has govern the occurrence of for delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

%Similar to the experience with the symmetric matter EoS, the next steps will be to connect these mean field to to observables and begin constraining them [cite danielewics science article and Cozma private communication] 
With the \spirit TPC, we were able to measure the pion yield without resorting to extrapolations, and were able to measure the pion energy spectra of low to high energy pions with high inefficiencies. In the following pages we will present the total pion yield and energy spectra of this dissertation, which complements the dissertaion of Jon Barney \cite{jon}.

%to provide much needed experimental data in which each code may compare too. The total theoretical uncertainty is too large relative to the experimental error bars to make a constraint on the density dependence of the symmetry energy. Any constraint would depend on the particular code one is using and would result in conflicting conclusions when using another theory. There is no good consensus at this time which codes should be favored over other codes, as the magnitude of the effects included in each code are still debated in order of importance. Here I will present the current level of agreement with the codes and remark on some of the implications it has moving forward.

@@

We begin with the total charged pion multiplicities for the two systems of extreme isospin asymmetry:  the neutron rich $\tin{132}{124}$ system and the neutron deficient  $\tin{108}{112}$ system. Though the previous FOPI data set provided significant constraints pertaining the the symmetry energy, they relied on extrapolations to low transverse momenta, $P_T = \sqrt{p_x^2 + p_y^2}$, below \SI{160}{\mega\electronvolt\per\clight} \cite{fopi}. In the \spirit TPC, we were able to measure pions to very low $P_T$ for rapidities $y > y{CM}$; where the rapidity is expressed along the beam direction (z) as,

@@

where $y_{beam}$ is the beam rapidity in the lab frame and $y_{CM}$ is the CM system rapidity. This is keeping with the same notation of $y_o$ in Reference \cite{fopi}. Here $y_o = 1$ corresponds to a particle moving at beam rapidity. The   $P_T - y_o$ phase space of the \spirit TPC is plotted for both charged pions in the $\tin{132}{124}$ system , in Figure~\ref{fig:ptrap_sn132}, and the $\tin{108}{112}$ system , in Figure~\ref{fig:ptrap_sn108}. Good efficiency down to very low $P_T$ avoided any need for extrapolations. There was good acceptance of pions $y_o > 0$ -- corresponding to $\theta_{CM} < \ang{90}$ -- where we cut off the lower rapidities. This marks the first time that pions have been measured to such low $P_T$.

@@

The integrated pion yield for both systems, and the $\pi^-/\pi^+$ ratio, is listed in Table~\ref{tb:pionyield}, where the systematic errors are the first error bar and the statistical error is listed next. Notice that the pion ratio is significantly greater than the $N/Z$ of the system.  This is not completely unexpected; in the Delta Resonance model where $\pi^-$ come primarily from neutron-neutron collisions and $\pi^+$ from proton-proton collisions, we expect  $\pi^-/\pi^+ = (N/Z)^2$ \cite{baoan_piprod1,baoan_piprod2}, where N/Z is the ratio of neutron to protons in the dense region where pions are produced. For the $\tin{132}{124}$ system $(N/Z)^2$ = 2.56 and for $\tin{108}{112}$ $(N/Z)^2$ = 1.44, but this naive scaling assumes no migration of nucleons in or out of the high density region where the isovector mean field and other transport properties may a impact that approximation significantly.

Due to this connection between pion ratio and the isospin asymmetry at high density, the pion ratio was proposed to be a good probe of the isovector dynamics at high density. was hypothesized to be proportional to the high density N/Z ratio of the early system, where the other effects such as pion absorption and re-emission would dilute the effect, lowering the pion ratio as the system tends toward isospin equilibrium. It is likely however, that the dynamics plays are more significant role in pion productions via the $\Delta(1232)$ resonance than the simple $(N/Z)^2$ relation, which was also seen in \cite{fopi}. Some codes assume the potential of the delta is just the same as the corresponding nucleon potential, as given by the iso-spin. While other codes -- namly TuQMD -- introduce $\Delta$ resonance in-medium potential, which has an iso-scalar component (independent of iso-spin) and an iso-vector component.  The nature of these two potentials is still unknown, and the role it plays in pion production has shown to be very important \cite{baoan_deltapotential, cozmaPC}.

@@

We have compared the total pion yields and ratios to the 7 common transport codes for the systems measured. The table of the transport codes are listed in Appendix~\ref{tb:pionyieldTheory}. Figure~\ref{fig:totalpiYield} shows the total pion yield for the four systems measured as compared with the codes. The codes plotted here are only the soft symmetry energy since the variation in code is much larger than the variation within a code between different symmetry energies. While some codes make a reasonable approximation of a particular charge pion yields, no code reasonably predicts both.

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{totalpiYield.png}
\caption{Total pion yields as compared with 7 common transport codes.}
\label{fig:totalpiYield}
\end{figure}

Figure~\ref{fig:totalpiRatio} shows the total single pion ratio, $R = Y(\pi^-)/Y(\pi^+)$, and the double ratio (DR), of the $\tin{132}{124}$ and $\tin{108}{112}$ system's single ratio, i.e. $DR = R_{\tin{132}{124}}/R_{\tin{124}{112}}$. Here, the variation between codes is much larger than the variation between the momentum dependence of the symmetry energy within a particular code. The symmetry energy variation (soft and stiff extremes) of two codes -- $\chi$BUU and TuQMD -- is plotted as a wide band in the single ratio and in all codes in the double ratio. The ciruclar markers and band represents the total error bar in the data. Certainly it can be seen that the variation between symmetry energy extremes in the codes, though small, still exists; as initially predicted \cite{baoan_piprod1,baoan_piprod2}, and the small error bars in the data would facilitate a detailed analysis to extract the high density behavior of the symmetry energy, if not for large disagreement between codes.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{totalpiRatio.png}
\caption{Total pion ratio and double ratio compared with 7 common transport codes.}
\label{fig:totalpiRatio}
\end{figure}

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the $N/Z=1.56$ where as in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.493$. We also expect the pion ratio is proportional to the $N/Z$ of the system. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Fig.~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. Scaling the other systems in the \spirit TPC data leads to almost the exact same value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

@@

Figures~\ref{fig:pionspectraSn132} and \ref{fig:pionspectraSn108} shows pion CM kinetic energy spectra for both the $\tin{132}{124}$ and $\tin{108}{112}$ systems respectively; corrected for efficiency and accounting for the solid angle of 4$\pi$. This data marks the first time pion spectral data has been measured at sub-threshold energies. In the pion spectra, we can see the effect of the coulomb potential, which accelerates $\pi^+$ and decelerates $\pi^-$ particles, due to the positive charge of the nuclear medium. The low $\pi^+$ production at low energy is caused by the coulomb barrier between protons which limits $\pi^+$ production at low energies. The Coulomb force has the largest effect on the spectra of the pions and will play an important role when observing the general shape of the pion spectral ratio. Besides the initial coulomb barrier for $\pi^+$ production, the coulomb force affects only the final energy of produced pions, and has almost no role in the total pion yield.

%Add figure on Glauber model 
%Add figure on comparison to FOPI data

@@

\begin{comment}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{singleRatio.png}
\caption{Single ratio spectra}
\label{fig:SRspectra}
\end{figure}
\end{comment}

@@

The pion spectral ratio is a rather promising observable. In particular, it may be more sensitive to the high density regions of the early collision. It is generally true, that high energy pions are more likely to exit the nuclear medium  earlier, and therefore be less prone to effects such as pion absorption and re-emission; all of which dilute the sensitivity of the pion observable to the high density behavior. Conversly, low energy pions are more likely to stay in the nuclear medium, and be affected by other effects such as the $\Delta$ potential in medium \cite{baoan_deltapotential}. Integrating these two regions to get the total pion yield dilutes the possible sensitivity the high energy pions have. Instead, we can construct the $Y(\pi^-)/Y(\pi^+)$ ratio as a function of the kinetic energy in the CM system with a particular focus on high energy pions.

@@

Figure~\ref{fig:SRsn132} and \ref{fig:SRsn108} shows the pion spectral ratio for the $\tin{132}{124}$ and $\tin{108}{112}$ system respectively, which was measured with a high degree of efficiency and accuracy. The general hyperbolic shape seen in both figures arises from the Coulomb force on the two charged pion spectra mentioned earlier in Section~\ref{sec:pionSpectra}. Also notice the pion ratio is smaller for less neutron poor system, as we would expect since less neutron-neutron collisions produce less $\pi^-$.

@@

In a similar way to the total pion double ratio, we can construct the spectral double ratio, the ratio between the two system's single ratios. In a similar way described in Section~\ref{sec:doubleRatio} we would expect systematic uncertainties in the experiment, and even in the theory --between the two systems-- to cancel out. For the same reasons as the pion spectral ratio, we would expect the high energy pions to be more sensitive to the high density region of the early collision. Here, just as in the total double ratio, the double ratio is significantly higher than any theory. Here as before, the bands in the theory represent the two extremes of the symmetry energy.

>>>>

 The data taken has been compared with 7 commonly used transport theoretical models. These models took part in a large collaborative effort in order to standardize certain common elements of each models \cite{theoryComp1,theoryComp2}. To reduce uncertainties in the pion production mechanism, each model simulated nuclear matter in a box type simulation with fixed, well defined, initial conditions. Though these box calculations model an equilibrated system at finite density, which is  not replicated by experiment, the solutions  of  pion and $\Delta$ yields can be analytically calculated from the ansatz of statistical equilibrium, providing a good benchmark. This allowed for each model to systematically go through the numerical treatment and details in each model such as initialization of nuclei, stability of the model, numerical handing of pauli-blocking, etc. After satisfying these benchmark calculations, each model was used with what each author considered the most reliable input and model assumption.  These models were taken in their best configuration, without any prior knowledge of the experimental data. We then simulated the 4 systems measured in the \spirit TPC at an impact parameter of \SI{3}{\femto\metre} at \SI{270}{\MeVA} beam energy. While the numerical treatments of each model are reasonably similar, each model differs considerably in their treatment of pion and $\Delta$ dynamics. Some models contain modifications to how the pion behaves in nuclear matter, i.e. \emph{in-medium effects}, typically introduced by including a pion optical potential, which describes the pion scattering and absorption. Some models include the iso-scalar and the iso-vector delta potential, which are not very well constrained, but can be important in the production of pions \cite{baoan_deltapotential,inmedPionKo,inmedPionFeng}.

%It is important to note that the interaction of pions and deltas with dense matter should not be viewed as a nasty detail that a "better" probe like nucleonic elliptical flow would avoid. 
One of the central questions regarding matter at twice saturation density concerns the importance of deltas and pions as important constituents of the matter at those densities \cite{pionNS,deltaNS,awayforward}. Questions of the role of deltas and pions at 2$\rho_o$ are strongly connected to these potentials. Calculations of pion production indicate that selected pionic observables such as the energy spectra and transverse momentum distributions can be used to constrain these potentials \cite{cozmaPC}. Obtaining such constraints are a major objective of our current work, but most of these constraints have not yet been obtained.

As will be the re-occurring theme, there is a large variation in the predicted pion observables between theoretical models; this is greater than the variation between the stiff and soft symmetry energies within a particular model. However, the remedy for this discrepancy is for the authors of these models to include or better model some other physics that many of them have neglected, or it could be that the relevant physics is not constrained by other observables and needs to be constrain by further experiments and comparisons.

For example, it is clear that many models need a  better description of the delta interactions with matter, which is critical to reproducing the total pion yields, and also the momentum dependence of the isovector mean field potentials, which have a large impact on the shapes of the energy spectra.

Other unconstrained physics includes aspects of the mean field potentials for nucleons, pions, and deltas. A primary focus is on the isovector potentials, of which only nucleon potentials for have been varied in the calculations shown below. Some of the models include the momentum dependent nucleonic mean field potentials, and the mean field potentials for deltas and pions. Calculations have already identified that the overall pion yields are influenced by the mean field potentials for the deltas \cite{cozmaPC}, which may not be surprising since a similar sensitivity has govern the occurrence of for delta and pionic matter within neutron stars \cite{deltaNS,pionNS}.

With the \spirit TPC, we were able to measure the pion yield without resorting to extrapolations, and were able to measure the pion energy spectra of low to high energy pions accurately. In the following pages we will present the total pion yield and energy spectra for the most asymmetric systems, which complements the measurements of $\tin{112}{124}$ and $\tin{124}{112}$ collisions, which were at the focus of the dissertation of Jon Barney \cite{jon}.

@@

We begin with the total charged pion multiplicities for the two systems of extreme isospin asymmetry: the neutron rich $\tin{132}{124}$ system and the neutron deficient  $\tin{108}{112}$ system. Though the previous FOPI data set provided interesting data relevant to  the symmetry energy, they relied on extrapolations to low transverse momenta, $P_T = \sqrt{p_x^2 + p_y^2}$, below \SI{100}{\mega\electronvolt\per\clight} \cite{fopi}. In the \spirit TPC, we were able to measure pions to very low $P_T$ for rapidities $y > y{CM}$; where the rapidity is expressed along the beam direction (z) as,

@@

where $y_{beam}$ is the beam rapidity in the lab frame and $y_{CM}$ is the CM system rapidity. This is keeping with the same notation of $y_o$ in Reference \cite{fopi}. Here $y_o = 1$ corresponds to a particle moving at beam rapidity. The   $P_T - y_o$ phase space of the \spirit TPC is plotted for both charged pions in the $\tin{132}{124}$ system , in Figure~\ref{fig:ptrap_sn132}, and the $\tin{108}{112}$ system , in Figure~\ref{fig:ptrap_sn108}. Good efficiency down to very low $P_T$ avoided any need for extrapolations. There was good acceptance of pions $y_o > 0$ -- corresponding to $\theta_{CM} < \ang{90}$ -- where we cut off the lower rapidities. This marks the first time that pions have been measured to such low $P_T$. We note that approximately 35\% of $\pi^+$ and 55\% of $\pi^-$ lie below the $P_T$ cut offs cited in the FOPI results \cite{fopi}. For the physics they were intending to do their extrapolations were sufficient, but for constraining the small effect of the symmetry energy, it is critical to measure all the way down to threshold for the pions to not introduce any extrapolation errors.

@@

The integrated pion yield for both systems, and the $\pi^-/\pi^+$ ratio, is listed in Table~\ref{tb:pionyield}, where the systematic errors are the first error bar and the statistical error is listed next. This represents data with an estimated average impact parameter of $b_{avg} = \SI{2.2}{\femto\metre}$ and $b_{max} < \SI{3}{\femto\metre}$. Notice that the pion ratio is significantly greater than the $N/Z$ of the system.  This is not completely unexpected; in the delta resonance model mentioned in Section~\ref{sec:pionObs}, we expect  $\pi^-/\pi^+ ~ (N/Z)^2$ \cite{baoan_piprod1,baoan_piprod2}, where N/Z is the ratio of neutron to protons in the dense region where pions are produced. For the $\tin{132}{124}$ system $(N/Z)^2$ = 2.56 and for $\tin{108}{112}$ $(N/Z)^2$ = 1.44, but this n\"aive model assumes no migration of nucleons in or out of the high density region where the isovector mean field and other transport properties may a impact that approximation significantly.

%Due to this connection between pion ratio and the isospin asymmetry at high density, the pion ratio was proposed to be a good probe of the isovector dynamics at high density. was hypothesized to be proportional to the high density N/Z ratio of the early system, where the other effects such as pion absorption and re-emission would dilute the effect, lowering the pion ratio as the system tends toward isospin equilibrium. It is likely however, that the dynamics plays are more significant role in pion productions via the $\Delta(1232)$ resonance than the simple $(N/Z)^2$ relation, which was also seen in \cite{fopi}. Some models assume the potential of the delta is just the same as the corresponding nucleon potential, as given by the iso-spin. While other models -- namly TuQMD -- introduce $\Delta$ resonance in-medium potential, which has an iso-scalar component (independent of iso-spin) and an iso-vector component.  These two potentials are poorly constrained, but the role they play in pion production in collisions and in neutron stars has been shown to be very important \cite{baoan_deltapotential, cozmaPC}.

@@

We have compared the total pion yields and ratios to the 7 common transport models for the systems measured. The table of the transport models are listed in Appendix~\ref{tb:pionyieldTheory}. Figure~\ref{fig:totalpiYield} shows the total pion yield for the four systems measured as compared with the models. The models plotted here are only the soft symmetry energy since the variation in model is much larger than the variation within a model between different symmetry energies. While some models make a reasonable approximation of a particular charge pion yields, no model reasonably predicts both.

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{totalpiYield.png}
\caption{Total pion yields as compared with 7 common transport models.}
\label{fig:totalpiYield}
\end{figure}

Figure~\ref{fig:totalpiRatio} shows the total single pion ratio, $R = Y(\pi^-)/Y(\pi^+)$, and the double ratio  $DR = R_{\tin{132}{124}}/R_{\tin{124}{112}}$, of the $\tin{132}{124}$ and $\tin{108}{112}$ systems. Here, the variation between models is much larger than the variation between the momentum dependence of the symmetry energy within a particular model. The symmetry energy variation (soft and stiff extremes) of two models -- $\chi$BUU and TuQMD -- is plotted as a wide band in the single ratio and in all models in the double ratio. The circular markers and band represents the total error bar in the data. Certainly it can be seen that the variation between symmetry energy extremes in the models, though small, still exists; as initially predicted \cite{baoan_piprod1,baoan_piprod2}, and the small error bars in the data would facilitate a detailed analysis to extract the high density behavior of the symmetry energy, if not for large disagreement between models.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{totalpiRatio.png}
\caption{Total pion ratio and double ratio compared with 7 common transport models.}
\label{fig:totalpiRatio}
\end{figure}

@@

The FOPI collaboration has measured the total pion multiplicity resulting from ${}^{197}Au + {}^{197}Au$ collisions at several, higher, beam energies. In the $\tin{132}{124}$ data the asymmetry is $N/Z=1.56$ where as in the ${}^{197}Au + {}^{197}Au$ the $N/Z=1.493$. Since 4 beams were measured in this experiment, the $N/Z$ dependence was measured as seen in Fig.~\ref{fig:totalpiRatio}. Here the dependence is fitted with a 2-nd order polynomial fit. To compare the pion ratio in the $\tin{132}{124}$ data with that of the lower N/Z in the FOPI experiment, we scaled by the pion ratio between $N/Z=1.56$ and 1.493 as given from the fitted polynomial line. Figure~\ref{fig:fopiPionRatio} shows the scaled pion ratio as compared with the FOPI pion ratio \cite{fopi}. The fitted function has the functional form of $p_1(E - p_2)^{-2}$ where $p_1$ and $p_2$ are free parameters, and only is meant to guide the eye. It is also worth mentioning that the pion ratio observed in the FOPI \SI{400}{\MeVA} setting was already considerably higher than what is expected from the $(N/Z)^2$ delta resonance model \cite{baoan_piprod1,baoan_piprod2}. It it worth mentioning that the two experiments differ considerably in their efficiencies as was mentioned before. The FOPI results heavily relied on extrapolations to lower pion energies which the \spirit TPC did not rely on \cite{fopi}. Without a systematic error calculation or discussion with the FOPI collaboration no further detail can be given except that the \spirit TPC data seems to be reasonable expectation and in general agreement with the FOPI data set. Scaling the other systems in the \spirit TPC data leads to almost the exact same scaled value of $\mathrm{Y}(\pi^-)/\mathrm{Y}(\pi^+) = \num{4.82}$ and would be redundant.

@@

Figures~\ref{fig:pionspectraSn132} and \ref{fig:pionspectraSn108} shows pion CM kinetic energy spectra for both the $\tin{132}{124}$ and $\tin{108}{112}$ systems respectively; corrected for efficiency and accounting for the solid angle of 4$\pi$. This data marks the first time pion spectral data has been measured at sub-threshold energies. In the pion spectra, we can see the effect of the Coulomb potential, which accelerates $\pi^+$ and decelerates $\pi^-$ particles, due to the positive charge of the nuclear medium. The low yield for the $\pi^+$ production at low energies is caused by the Coulomb barrier between protons which limits $\pi^+$ production at low energies. The Coulomb force has the largest effect on the spectra of the pions and will play an important role when observing the general shape of the pion spectral ratio, where it changes its shape completely.

@@

@@

The pion spectral ratio is a rather promising observable. In particular, it may be more sensitive to the high density regions of the early collision. It is generally true that high energy pions are more likely to exit the nuclear medium  earlier, and therefore be less prone to effects such as pion absorption and re-emission; all of which dilute the sensitivity of the pion observable to the high density behavior. Conversely, low energy pions are more likely to stay in the nuclear medium, and be affected by other effects such as the $\Delta$ potential in medium \cite{baoan_deltapotential}. Integrating these two regions to get the total pion yield dilutes the possible sensitivity the high energy pions have. Instead, we can construct the $Y(\pi^-)/Y(\pi^+)$ ratio as a function of the kinetic energy in the CM system with a particular focus on high energy pions.

@@

Figure~\ref{fig:SRsn132} and \ref{fig:SRsn108} shows the pion spectral ratio for the $\tin{132}{124}$ and $\tin{108}{112}$ system respectively, which was measured with a high degree of efficiency and accuracy. The general hyperbolic shape seen in both figures arises from the Coulomb force on the two charged pion spectra mentioned earlier in Section~\ref{sec:pionSpectra}. Also notice the pion ratio is smaller for the neutron poor system, as we would expect since there are fewer neutron-neutron inelastic collisions and consequently fewer $\pi^-$.

@@

In a similar way to the total pion double ratio, we can construct the spectral double ratio, the ratio between the two system's single ratios. As described in Section~\ref{sec:doubleRatio} we would expect systematic uncertainties in the experiment, and even in the theory --between the two systems-- to cancel out. For the same reasons as the pion spectral ratio, we would expect the high energy pions to be more sensitive to the high density region of the early collision. Here, just as in the total double ratio, the spectral double ratio is significantly higher than any theory. Here as before, the bands in the theory represent the two extremes of the symmetry energy.

### 2020-07-17 chapters/ch_summary.tex
Added bills edits working on appendix figures

\chapter{Summary}
The need for precise experimental pion data is paramount to understanding exotic, dense matter, such as neutron stars. Besides direct observation of these objects, the lab is the best way to probe the density dependence of the symmetry energy, and is the only way to probe different asymmetries. Pions have been proposed to be sensitive to the high density regions of these collisions, making it an ideal probe to constrain the symmetry energy. Though simulating pion production has proved to be quite challenging, it has motivated the design of high efficiency detectors such as the \spirit TPC to measure pion emission from neutron rich heavy ion collisions. 
We preformed a campaign of experiments measuring collisions of neutron rich, radioactive beams, on stable targets in 4 different configurations -- $\tin{132}{124}, \tin{108}{112}, \tin{112}{124},$ and $\tin{124}{112}$. All measured with the \spirit TPC which was able to measure pions to very low energies without the need for any extrapolations and very little assumptions.

\subsection{Results}
The pion yield, ratio, and some new observables -- namely the double ratio-- were measured to within <4\% accuracy. We observed a deviation from the  naive $\pi^-/\pi^+ ~ (N/Z)^2$ scaling expected from the $\Delta$-resonance model discussed in Section~\ref{sec:pionObs}.

\subsection{Outlook}

>>>>

\chapter{Summary}
The need for precise experimental constraints on the symmetry energy at high densities is paramount to understanding exotic, dense matter, such as neutron stars. Direct observation of these objects provides information about the total equation of state, which is primarily sensitive to the pressure in neutron matter. Only by laboratory measurements can one control the asymmetry of matter and isolate its dependence on the isospin asymmetry $\delta = \frac{\rho_n - \rho_p}{\rho}$. This information is critical to understanding the symmetry energy and associated isovector mean field potentials. These mean field potential contribute strongly to the chemical potentials, which control the phase transitions and the internal structure within stars. Heavy ion collisions provide the only laboratory environment to study these issues at the suprasaturation densities relevant to the neutron star interior.

Pions have been proposed to be sensitive to the high density regions of these collisions, making it a compelling probe to constrain the symmetry energy. Moreover, questions about what are the stable phases at twice saturation densities cannot be answered without understanding the mean field potentials for pions and deltas in such matter. Though simulating pion production has its challenges, the prospect of answering such compelling questions has motivated the design of high efficiency detectors such as the \spirit TPC to measure pion emission from neutron rich heavy ion collisions. We have preformed a campaign of experiments measuring collisions of neutron rich, radioactive beams, on stable targets in 4 different configurations -- $\tin{132}{124}, \tin{108}{112}, \tin{112}{124},$ and $\tin{124}{112}$, all measured with the \spirit TPC.

\subsection{Results}
The pion yield, ratio, and some new observables -- namely the double ratio-- were measured to within <4\% accuracy. We find that a major fraction of the pions are emitted below the thresholds of the FOPI experiment, demonstrating that the systematic errors are of the pion ratios from that previous experiments are completely dominated by the extrapolation of the FOPI data to the emission threshold. Based on this information, it is clear that prior conclusions based on comparison of those extrapolated data to theoretical calculations of total pion yields do not take the systematic errors of that extrapolation into account and any agreement their conclusions would contain error but to an unknown degree.

  This marks the first time pions have been accurately measured in the sub-threshold region. We observed significant a deviation from the  n\"aive expectation of the simple $\Delta$ resonance model, where $\pi^-/\pi^+ ~ (N/Z)^2$, as discussed in Section~\ref{sec:pionObs}. Such a deviation is also predicted by theory, so this explanation should be retired in favor of something more quantitative and informative. Obtaining a better explanation from current calculations is difficult because nearly all pion production calculations are incomplete and each model has a mixture of strengths and inadequacies that are unique to each model. Thus, it is difficult to get the theorists to agree about the consequences of some of the model assumptions that they employ; and consequently a constraint on the density dependence of the symmetry energy.

  %With this wealth of new data that we have obtained, we now begin to work with theorists to understand what new experimental pion observable is sensitive to which theoretical assumption. We will are exploring how the isoscalar and isovector $\Delta$ potential plays an important role in the pion production spectra and emission rates. 
  Also for the first time, the pion spectra was measured. The spectral ratio appears to provide information that is relevant to this question also providing much needed information about the dynamics of pions, even to very low pion energies. Of particular importance is the sensitivity of calculations to the high energy tails of the pion spectra. This is a promising observable since high energy pions  exit the nuclear matter sooner and are less affected by re-scattering and adsorption-re-emission processes, which dilute information about the EOS at the highest density. We expect that high energy pions will be more sensitive to the prevailing conditions at high density, while low energy pions are more sensitive to the energy available for their emission, such as the Coulomb potential, the pion optical potentials, and the $\Delta$ potential.

\subsection{Outlook}
Theoretical collaboration, which has made considerable progress improving details of the calculation, will further improve in light of this new data. It is clear that a serious effort must be made to include effects such as the symmetry potentials of pions and deltas; which most codes do not. It also appears to be possible to make first constraints on the generally unknown $\Delta$ potential by fitting the pion yield data \cite{cozmaPC}. Making such a constraint would allow for the possibility to start making constraints on the density dependence of the symmetry energy. The interactions with theoretical community that has coalesced around this effort is very exciting, and we are seeing real progress in transport theory. It seems very possible that the theoretical efforts will converge and provide solid interpretations of these and other data much more quickly than they did in the case of the symmetric matter EoS.
